---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clusterfuck

<!-- badges: start -->
<!-- badges: end -->

Make clustering data sets easy with `'clusterfuck'`!
Use it to categorize data into clusters, choose representative data points and broad structures in the data.

A main feature of this package is the clustering of **non numeric** or **mixed** data and clustering using exotic **custom distance functions**.

The implemented clustering algorithms:

 - K-Means
 - K-Medoids
 - spectral clustering
 - hierarchical clustering
 - DBSCAN
 - OPTICS


Additionally it includes

- functions for **generating** Data to test clustering functions on
- a collection of pre made **data for clustering**
- a collection of pre made **distance functions** for clustering (some very exotic)
- functions to **view data** and **view clustered data** of up to 3 Dimensions
- a function to calculate the **silhouette coefficient** of a clustering

## Installation

You can install the development version of clusterfuck from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("adamfuge2/R-Projekt-Haufenbildung")
```

## Examples

Lets load clusterfuck
```{r example}
library(clusterfuck)
```

### Clustering some points in 2D

Lets take a look at a data set
```{r example_1}
euclidean_cluster_data |> viewData()
```

(euclidean_cluster_data is one of the premade data sets and viewData a function to view it)

We would like to cluster this data set into three distinct parts. We can use K-Means to do that.

```{r cluster_euclid}
kMeans(euclidean_cluster_data, K=3)
```

The result is a clustering object with some more useful information. 
See the vignette of kMeans for more information about it.

### Clustering exotic data sets and distances
An exotic example data set featured in `'clusterfuck'` is
```{r}
study_courses_data
```

with its (made up) distance function `'study_courses_distance'`. 
Because is a main feature of the package it is very easy to cluster this pair of data and distance function. 

```{r}
hierarchicalClustering(study_courses_data, 
                       custom_distance_function = study_courses_distance)
```

### Clustering misshapen data without custom distance
K-Means and K-Medoids work very well when the clusters are **spherical** meaning in a filled 'round' ball around a center.

Lets look at this data set.
```{r, results='hide'}
two_concentric_circles |> viewData()
```

K-Medoids and K-Means have a very hard time clustering this
```{r, results='hide'}
kMedoids(two_concentric_circles,K=2)
```

We could now invest the effort and try to define a distance function on this data to improve the clustering made by K-Medoids and K-Means (usually one would choose K-Medoids for these kinds of distance functions).

But far simpler is the approach to use DBSCAN or OPTICS:
```{r dbscan, results='hide'}
clustering <- optics(two_concentric_circles,epsilon = 0.3,min_Pts = 5)
```

```{r dbscan_visualize, results='hide', echo=FALSE}
# For visualization purposes
clustered_data(two_concentric_circles,
               extractClusters(clustering,epsilon=0.3)) |>
  viewClusters()
```

### Clustering is hard
Now lets look at a data set that's famously hard to cluster:
```{r, results='hide'}
three_connected_concentric_circles |> viewData()
```

K-Means and K-Medoids fail because the data is non spherical
```{r three_con_kmeans, results='hide'}
kMeans(three_connected_concentric_circles, K=3)
```

DBSCAN, OPTICS and hierarchical clustering fail because the data points are all connected
```{r}
dbscan(three_connected_concentric_circles,epsilon = 0.1,min_Pts = 5)
```

Our last resort: the powerful spectral clustering:
```{r}
spectralClustering(three_connected_concentric_circles,
                   k = 3,
                   gamma = 50,
                   cluster_algorithm = 'K-Medoids',
                   K = 3)
```

**Success!**


