---
title: "Data for clustering"
output:
  rmarkdown::html_vignette:
    toc: true
    tabset: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Data for clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 7,
  dev='svg'
)
```

```{r setup}
library(clusterfuck)
```

The clusterfuck package provides a generous amount of data to test the clusterings on, as well as some generators to test it. We will first explore these Generators and then take a look at the pre made data sets

# Data Generators
If you want to generate some Data for testing clustering functions, start by using the powerful and versatile [generateClusterData](#generateClusterData) and if that does not suffice check out the other generators. 

## generateClusterData
This function generates data perfect for clustering. 
Note that this will generate data in **spherical** clusters.

The Parameters:

- `n`           A positive integer. The number of total data points to be
 generated.
- `cluster_amount`  A positive integer. The number of clusters to
 generate. If none given, choose a random amount less than both $2^{dim}$ and $\frac{n}{4}$.
- `dim` A positive integer. The dimension of the data points to be
 generated
- `lower_bounds`,`upper_bounds` numeric vectors with length equal dim (if
 given). Bound the cluster centers in a \code{dim}-dimensional rectangle defined
 by these parameters. Beware: These bound the cluster centers and **not**
 the data points!
- `clusters_mean` A tibble with every row representing a center of a
 cluster. The column amount must match the dimensions implied by `dim` or the
 bounds.
- `clusters_sd` A numeric vector with length matching the \code{cluster_amount}
 or number of rows in `clusters_mean`. The standard deviations used in the
 generation of data points around the cluster centers (using a symmetrical
 normal distribution).
- `clusters_prob` A numeric vector with length matching the cluster_amount
 or number of rows in `clusters_mean`. The probability weights used to choose
 which which cluster a generated data point belongs to.
- `colnames` A character vector with length matching dim or the length of
 the bounds. Default is `'X_1'` to `'X_n'`
- `include_cluster` A logical of length 1. If `TRUE` a column `'cluster'`
 will be included in the generated data which will also be assigned the class `'clustered_data'`.
- `.print_info` A logical of length 1. If `TRUE` additional information
 will be displayed during runtime. Used in debugging.
 
 Examples:
```{r generateClusterData_examples}
generateClusterData(cluster_amount = 4)|> viewClusters()
generateClusterData(n = 1000,
                    clusters_mean =
                       tibble::tibble(longitude = c(-100,-60,10,20,70,130),
                                      latitude =  c(  40,-10,50,20,50,-20)),
                    clusters_sd = c(8,8,3,6,15,5),
                    clusters_prob = c(20,20,10,100,100,10),
                    include_cluster = TRUE) |> viewClusters()
generateClusterData(n = 1,
                    lower_bounds = -(20:1),
                    upper_bounds = 1:20,
                    dim = 20,
                    .print_info = TRUE) 
```

## generateClusterDataFromPaths
If you want to generate clusters that have **non spherical** shapes use this function.
It generates data points along paths.

The parameters:
- `n`               a positive integer. The number of total data points to
  be generated.
- `list_of_paths`   a list containing tibbles. Each tibble containing
  points (in rows) to be interpreted as paths along which the data is
  generated.
- `clusters_sd` A numeric vector with the same length as `list_of_paths.`
  The standard deviations used in the generation of data points around the
  clusters (using a symmetrical normal distribution).
- `clusters_prob` A numeric vector with the same length as `list_of_paths`.
  The probability weights used to choose which which cluster a generated data
  point belongs to.
- `include_cluster` A logical of length 1. If `TRUE` a column `'cluster'`
  will be included in the generated data which will also be assigned the class `'clustered_data'`.
- `.print_info` A logical of length 1. If `TRUE` additional information
  will be displayed during runtime. Used in debugging.

Some Examples:
```{r}
generateClusterDataFromPaths(list_of_paths =
                                list(tibble::tibble(X = c(0,1,1),
                                                    Y = c(0,0,1)),
                                     tibble::tibble(X = c(0,0.8),
                                                    Y = c(1,0.2))))|> viewClusters()
generateClusterDataFromPaths(n = 1000,
                             list_of_paths =
                                list(tibble::tibble(X = c(0),
                                                    Y = c(0)),
                                     tibble::tibble(X = c(1),
                                                    Y = c(1))),
                             clusters_sd = c(0.1,0.5),
                             clusters_prob = c(1,2),
                             include_cluster = TRUE,
                             .print_info = TRUE) |> viewClusters()

```

## generateNoiseData
In the generators above the data points will always be clearly assigned to clusters. 
This does not always reflect real world data. 
generateNoiseData generates a rectangular space filled uniformly with data points. If you include the cluster of each data point they will be all assigned to the Cluster 0 - the 'Noise' Cluster.


- `n`           A positive integer. The number of total data points to be
 generated.
- `lower_bounds`,`upper_bounds` numeric vectors with equal length. Bounds the (multidimensional) rectangle in which the data points will be generated.
- `colnames` A character vector with length matching dim or the length of
 the bounds. Default is `'X_1'` to `'X_n'`.
- `include_cluster` A logical of length 1. If `Noise` a column
 `'cluster'` will be included in the generated data which will also be
 assigned the class `'clustered_data'`. Because this is a noise generator
 all data points will be assigned cluster 0.
- `.print_info` A logical of length 1. If `TRUE` additional
 information will be displayed during runtime. Used in debugging.
 
```{r generateNoiseData}
generateNoiseData(lower_bounds = c(-1,0),
                 upper_bounds = c(1,1))
generateNoiseData(n = 1000,
                 lower_bounds = c(-180,-90),
                 upper_bounds = c(180,90),
                 colnames = c('longitude','latitude'),
                 .print_info = TRUE)

```


# Data sets included in clusterfuck
There are quite a few nice example data sets included in this package.
We will go over them here and, if relevant, describe how different clustering algorithms act on it.

Note: During this vignette we will make heavy use of the function `viewData` and `viewClusters` which are just nice little wrappers of plotting functions and should not distract you from the data sets.

## optics_wikipedia_data.
Imitates the Data set featured on the Wikipedia page for OPTICS.

```{r}
optics_wikipedia_data |> viewData()
```

## euclidean_cluster_data

```{r}
euclidean_cluster_data |> viewData()
```

We also have the same data set with clusters included

```{r}
euclidean_clustered_data |> viewClusters()
```

## manhattan_cluster_data
This data set should be clustered using manhattan distances.

```{r}
manhattan_cluster_data |> viewData()
```


We also have the same data set with clusters included


```{r}
manhattan_clustered_data |> viewClusters()
```

## maximum_cluster_data
This data set should be clustered using maximum (norm) distances.

```{r}
maximum_cluster_data |> viewData()
```


We also have the same data set with clusters included


```{r}
maximum_clustered_data |> viewClusters()
```

## concentric circles
These data sets make K-Means and K-Medoids **fail**. 
(If not using any fancy distance functions)

### two_concentric_circles
```{r}
two_concentric_circles |> viewData()
```

### three_concentric_circles

```{r}
three_concentric_circles |> viewData()
```

## connected concentric circles
These data sets make

- K-Means and K-Medoids using standard distance functions **fail** 
-  [hierarchical clustering], [optics()] and [dbscan()] **fail** because they would all find one big connected cluster

We suggest using spectral clustering on this one
 
### two_connected_concentric_circles 
 
```{r}
two_connected_concentric_circles |> viewData()
```

### three_connected_concentric_circles

```{r}
three_connected_concentric_circles |> viewData()
```

## looping data sets
The following are data sets are data sets in 'pacman' (go out to the right, come in from the left) space embedded in $\mathbb{R}^2$.


### hours_of_the_day
```{r}
hours_of_a_day |> viewData()
```

### hours_full_data

```{r}
hours_full_data |> viewData()
```

### hours_data

```{r}
hours_data |> viewData()
```

### oh_no_ive_spilled_my_pacman_dots
Not: This is a nice data set with data points sharing the same exact coordinates

```{r}
oh_no_ive_spilled_my_pacman_dots |> viewData()
```

### pacman_full_data

```{r}
pacman_full_data |> viewData()
```

## study_courses_data
This is a data set with char values.

- As `mean()` is ill defined on characters, K-Means cannot be used on it.

```{r}
study_courses_data |> viewData()
```

## numbers_and_characters
A data set with a column of type double and a column of type character.
Must be accompanied by applicable distance function when clustering. 

```{r}
numbers_and_characters |> viewData()
```
