---
title: "The DBSCAN Clustering Algorithm"
output: 
  rmarkdown::html_vignette:
    toc: true
    tabset: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The DBSCAN Clustering Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 7,
  dev='svg'
)
```

```{r setup, results=FALSE , message=FALSE, warning=FALSE}
library(clusterfuck)
```

This is an overview about the DBSCAN (Density-Based Spatial Clustering of Applications with Noise) Clustering algorithm of the clusterfuck package.
The algorithm works in a way that a cluster is formed if a specified number of points min_points is within a specified radius epsilon to each other. 

# General use
We start with creating some test data
```{r general_use_1}
data <- generateClusterData(n=300,clusters_sd = c(0.04,0.01,0.05,0.04,0.04,0.04))
```

This is what it looks like
```{r general_use_2}
viewData(data)
```

**Apply the DBSCAN algorithm with a neighborhood of 0.2 and a minimal number of points for a cluster to be formed 3**
```{r general_use_3}
result <- dbscan(data, epsilon=0.2, min_Pts=3)
result
```

# Parameters

## data (required)
This is the main parameter to be provided and the data upon which the clustering algorithm will be deployed. It is recommended your data is a tibble with every row representing a data point. Must have at least 1 data point having at least 1 dimension. 

Here some examples: 
```{r data_examples_1}
data_1 <- generateClusterData(n=100)
data_2 <- generateClusterData(n=2000)
data_3 <- generateClusterData(n=1)
```

We can cluster this, and visualize the result:
```{r data_examples_2}
result_1 <- dbscan(data_1, epsilon=0.2, min_Pts=3)
viewClusters(result_1$clustered_data)
result_2 <- dbscan(data_2, epsilon=0.2, min_Pts=3)
viewClusters(result_2$clustered_data)
result_3 <- dbscan(data_3, epsilon=0.2, min_Pts=3)
viewClusters(result_3$clustered_data)
```
Points assigned to cluster 0 represent noise points that were not
assigned to any cluster.

## epsilon (required) {.tabset}
The neighborhood radius from a point in which other points are considered neighbors.
epsilon must be a positive number greater than 0. The larger epsilon is set, the less clusters will be formed, as more and more points will fall into the same cluster. A smaller epsilon on the other hand will lead to more clusters.

This can be be observed when clustering the same data with different epsilon values:

### small epsilon
```{r epsilon_examples_1a}
result_small_epsilon <- dbscan(optics_wikipedia_data, epsilon=0.015, min_Pts=5)
viewClusters(result_small_epsilon$clustered_data)
```

### medium epsilon
```{r epsilon_examples_1b}
result_medium_epsilon <- dbscan(optics_wikipedia_data, epsilon=0.1, min_Pts=5)
viewClusters(result_medium_epsilon$clustered_data)
```

### high epsilon
```{r epsilon_examples_1c}
result_big_epsilon <- dbscan(optics_wikipedia_data, epsilon=0.3, min_Pts=5)
viewClusters(result_big_epsilon$clustered_data)
```

## min_Pts (required) {.tabset}
The number of points that have to be within epsilon neighborhood of each other to count as a cluster.
min_Pts must be a positive integer greater than 0. The larger min_Pts is set, the less clusters will be formed, as less clusters have enough points within epsilon neighborhood. A small min_Pts will lead to more clusters.
Large min_Pts will lead to more noise points.

This can be be observed when clustering the same data with different min_Pts values:

### small minPts
```{r minPts_examples_1a}
result_small_minPts <- dbscan(optics_wikipedia_data, epsilon=0.05, min_Pts=2)
viewClusters(result_small_minPts$clustered_data)
```


### medium minPts
```{r minPts_examples_1b}
result_medium_minPts <- dbscan(optics_wikipedia_data, epsilon=0.05, min_Pts=10)
viewClusters(result_medium_minPts$clustered_data)
```


### high minPts
```{r minPts_examples_1c}
result_big_minPts <- dbscan(optics_wikipedia_data, epsilon=0.05, min_Pts=100)
viewClusters(result_big_minPts$clustered_data)
```

## distance_method, p, custom_distance_function (optional)
These parameters control how distances between data points are computed.
They behave identical to the parameters described in the K-Means vignette.

# Output
The output is a clustering object:
A list of the class "clustering" with two elements clustered_data and clustering_function.

## clustered_data
The original data with an additional column called "cluster" added.
```{r clustered_data}
result$clustered_data
```

## clustering_function
A function which can relate any datapoint to their cluster.
```{r clustering_function_example}
clusters <- list(
  tibble::tibble(X = c(0.2,0.3,0.4,0.3,0.2), Y = c(0.3,0.4,0.3,0.2,0.3)),
  tibble::tibble(X = c(0.45,0.5,0.45,0.4,0.45), Y = c(0.9,0.85,0.8,0.85,0.9)),
  tibble::tibble(X = c(0.85,0.8,0.93,0.85), Y = c(0.5,0.28,0.28,0.5))
)

training_data <- generateClusterDataFromPaths(n = 200, clusters)
more_data  <- generateNoiseData(n = 500, lower_bounds = c(0,0), upper_bounds = c(1,1))

# derive clustering using DBSCAN
f <- dbscan(training_data, epsilon = 0.15, min_Pts = 3)$clustering_function

f
f(c(0.5,0.5))

# The clustering function can now be applied to different datasets.
viewClusters(training_data, f)
viewClusters(more_data, f)
```



