---
title: "The Hierarchical Clustering Algorithm"
output: 
  rmarkdown::html_vignette:
    toc: true
    tabset: true
vignette: >
  %\VignetteIndexEntry{The Hierarchical Clustering Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 7,
  dev = 'svg'
)
```

```{r setup, message = FALSE, warning = FALSE}
library(clusterfuck)
```

This is an overview about the Hierarchical Clustering algorithm of the clusterfuck package.

## General Use
In the most simple case, we have a 2-dimensional dataset containing, for example, 4 clusters:

```{r general_use_1, echo = 2}
set.seed(129)
data <- generateClusterData(n = 100, cluster_amount = 4, dim = 2)
```

This data set can be illustrated by using the \code{viewData} function:

```{r general_use_2}
viewData(data)
```

**Apply Hierarchical Clustering with 4 Clusters:**

```{r general_use_3}
hierarchicalClustering(data, 4)
```
This returns a \code{clustering} object.


## Parameters

### 2.1 data (necessary)
This is the main parameter to be provided and the data upon which the clustering algorithm will be deployed. It is recommended your data is a tibble with every row representing a data point or alternatively a matrix. 

Some examples:
```{r data_definitions, echo = c(2,3)}
set.seed(121)
data_1 <- generateClusterData(n = 100, cluster_amount = 7)
data_2 <- generateClusterData(n = 1)
```

These can be clustered by the standard usage of Hierarchical Clustering:

```{r standard_usage}
hierarchicalClustering(data_1, 3)
hierarchicalClustering(data_2, exact_cluster_amount=1)
```



### 2.2 K
The number of clusters the algorithm tries to fit the data in.
This number must be
- a positive integer 
- larger than 0
- at maximum equal to the amount of unique data points in `data`

Some examples
```{r K_examples}
hierarchicalClustering(data_1, 3)
hierarchicalClustering(data_1, 5)
hierarchicalClustering(data_1, 10)
hierarchicalClustering(data_2, exact_cluster_amount = 1)
# hierarchicalClustering(data_2, 5) throws error
```


## 2.3 mode
The linkage mode used to calculate the distance between two clusters. Since, in 
general, a cluster consists of multiple datapoints, the definition of a distance
between two clusters cannot be deduced by the distance function ('metric').

To tackle this problem we can introduce methods for employing distance functions
defining a distance between clusters. 

These modes can be transferred as an optional argument in the form of a string.
In the following, the usage of these modes is demonstrated.

### 2.3.1 Centroid
One rather heuristic method to define intercluster distance is to determine the 
centroids of both clusters (this process is independet of any distance function
as it makes use of the arithmetic mean) and then calculating the distance between
the two centroids with the distance function.

```{r mode-options-centroid, eval = FALSE}
hierarchicalClustering(data, 4, mode = "centroid")
```

### 2.3.2 Single
Interprets the smallest distance of the pair \code{p_1, p_2} with \code{p_1} belonging
to cluster 1 and \code{p_2} to cluster 2 as the distance beween the two clusters.

```{r mode-options-single, eval = FALSE}
hierarchicalClustering(data, 4, mode = "single")
```

### 2.3.3 Complete
Interprets the largest distance of the pair \code{p_1, p_2} with \code{p_1} belonging
to cluster 1 and \code{p_2} to cluster 2 as the distance between the two clusters.

```{r mode-options-complete, eval = FALSE}
hierarchicalClustering(data, 4, mode = "complete")
```


### 2.3.4 Average
Calculates the distance between every pair \code{p_1, p_2} with \code{p_1} belonging
to cluster 1 and \code{p_2} to cluster 2 and then returning the mean of these values.

```{r mode-options-average, eval = FALSE}
hierarchicalClustering(data, 4, mode = "average")
```




### 2.4 distance_method
This function comes with a handful of common distance function ('metric') pre made. 
Use this parameter to choose one of the following:

- `'euclidean'` (default)
- `'maximum'`
- `'manhattan'`
- `'Lp'`


#### Examples {.tabset}

##### euclidean
```{r metric_examples_euclidean}
hierarchicalClustering(euclidean_cluster_data, exact_cluster_amount =2, distance_method = 'euclidean')
```

##### maximum
```{r metric_examples_maximum}
hierarchicalClustering(maximum_cluster_data, exact_cluster_amount=2, distance_method = 'maximum')
```

##### manhattan
```{r metric_examples_manhattan}
hierarchicalClustering(manhattan_cluster_data, exact_cluster_amount=2, distance_method = 'manhattan')
```


### 2.5 p
If you chose `'Lp'` for the parameter `distance_method` then this is where you can input the value for the p-metric. Will be ignored, if other distance functions are chosen.



### 2.6 custom_distance_function
You can also choose a custom distance function. This distance function must be a function with the following properties. If a custom distance function is provided, the parameters `distance_method` and `p` are ignored. 

But please consider:
- **Using an unoptimized custom distance function can slow** the calculation of clustering algorithms down significantly! 

Inputs:   
- `x` and `y` : Two of the data row type. (The internally chosen names do not matter)

Output: 
- Must be greater or equal to 0 (semi-definite)
- Must be zero, if `x` and `y` are identical
- Must have the same output, when the values of `x` and `y` are switched. (symmetry)


Examples:

```{r globe_example_distance}
# the custom function for distance on a globe
distanceByLongitudeAndLatitude <- function(x,y){
  x <- unlist(x,use.names = FALSE)[2:1] * 2*pi/360
  y <- unlist(y,use.names = FALSE)[2:1] * 2*pi/360
  round(2*6371000 * asin(sqrt(sin((y[1]-x[1])/2)^2 +
                       cos(x[1]) * cos(y[1]) *
                     sin((y[2]-x[2])/2)^2)),2)
}
```

```{r globe_example_1, echo = -3}
if(require('maps')){
# load some data of word cities
set.seed(124)
slice_of_the_world <- tibble::tibble(dplyr::slice_sample(n=200, maps::world.cities[ ,c('long','lat')]))

hierarchicalClustering(slice_of_the_world, custom_distance_function = distanceByLongitudeAndLatitude)
} 
```

Because this is a main feature of the package, here is another example:


```{r custom_distance_function_example_2}
WIDE_euclidean <- function(x,y) {
  x[[1]] <- 0.1*x[[1]]
  y[[1]] <- 0.1*y[[1]]
  euclidean(x,y)
  }
hierarchicalClustering(data_1, min_cluster_amount =4, custom_distance_function = WIDE_euclidean)
```


### 2.7 .print_info
A logical of length 1.
This is useful for debugging or finding loops caused by custom metrics. If `TRUE` some extra info will be displayed in the console.

Example: 
```{r .print_info_example}
hierarchicalClustering(data_1, .print_info = TRUE)
```




## 3 Output
The output is a clustering object:
A list of the class clustering with at least an element called `'clustered_data'`.

### 3.1. clustered_data
The original data with a column called `'cluster'` added.

```{r clustered_data}
hierarchicalClustering(data_1, exact_cluster_amount =4)$clustered_data
```

### 3.2 cluster_amount
The amount of clusters deferred by the algorithm

```{r cluster_amount}
hierarchicalClustering(data_1)$cluster_amount
```

