---
title: "The K-Medoids Clustering Algorithm"
output: 
  rmarkdown::html_vignette:
    toc: true
    tabset: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The K-Medoids Clustering Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 7,
  dev='svg'
)
```

```{r setup, results = FALSE}
library(clusterfuck)
```

Here we will go over all information needed to use the kMedoids function of the clusterfuck package

# General use

## Base case
This will illustrate the most basic use case for this clustering function:

Lets create some test data
```{r general_use_1}
data <- generateClusterData(n=100, cluster_amount = 4)
```

This is what it looks like
```{r general_use_2}
viewData(data)
```

**Apply the K-Medoids-algorithm with K = 4**       
```{r general_use_3}
kMedoids(data, K = 4)
```

## How to find the right cluster amount
K-Medoids can be used to find the right cluster amount. 
Here we will use the **in clusterfuck implemented** silhouette coefficient.
The clustering object already includes a value for the [mean_silhouette](#mean_silhouette) of the clusterings. High values of the mean silhouette imply a very good fit.


```{r}
findClusterAmountSilhouette <- function(data,distance_method='euclidean',p=NULL,custom_distance_function=NULL ,check_min = 1){
  clusterings <- list()
  fit <- list()
  improvement <- Inf
  K <- 1


  while(improvement>0 || K <= check_min){

    fit[[K]] <- kMedoids(data = data,
                          K = K,
                          distance_method = distance_method,                          
                          p=p,
                          custom_distance_function=custom_distance_function
                          )$mean_silhouette

    if(K>1) improvement <- fit[[K]] - fit[[K-1]]

    K <- K+1

  }


  plot(1:(K-1),fit)

  return(which.max(fit))
}
```

Lets try it:
```{r}
findClusterAmountSilhouette(study_courses_data, 
                            custom_distance_function = study_courses_distance,
                            check_min = 10)
```




# Parameters

## data
This is the main parameter to be provided and the data upon which the clustering algorithm will be deployed. It is recommended your data is a tibble with every row representing a data point. Must have at least 1 data point. 

Here some examples: 
```{r data_examples_1}
data_1 <- generateClusterData(n=100)
data_2 <- generateClusterData(n=1)
data_3 <- study_courses_data
data_5 <- generateClusterData(n=100,
                              clusters_mean = tibble::tibble('X'=c(0.4,0.4,0.5,1),
                                                             'Y'=c(0.7,0.1,0.5,1)),
                              clusters_sd=c(0.01,0.02,0.01,0.05),
                              cluster_amount = 4)
```

The data can even have multiple types
```{r data_examples_2}
numbers_and_characters
```

```{r data_examples_3, include=FALSE}
numericAndCharacterMixedDistance <- function(x,y){
 return(length(x[[2]])-length(y[[2]]) +  abs(x[[1]]-y[[1]]))
}
```


Lets go clustering!
```{r data_examples_4, results = FALSE}
kMedoids(data_5,K=3)
kMedoids(data_1,K=1,.print_info = TRUE)
kMedoids(data_3,K=3,custom_distance_function = study_courses_distance)$clustered_data
kMedoids(numbers_and_characters,K=3,custom_distance_function = numericAndCharacterMixedDistance)
```

## K
The number of Clusters the Algorithm tries to fit the data in.
This number must be
- a positive integer 
- larger than 0
- at maximum equal to the amount of unique data points in data

Some examples:
```{r K_examples, results = FALSE}
kMedoids(data_1,K=3)
kMedoids(data_1,K=5)
kMedoids(data_1,K=10)
kMedoids(data_2,K=1)
```

## distance_method {.tabset}
This function comes with a handful of common distance functions pre made. Use this parameter to choose one of the following:

- 'euclidean' (default)
- 'maximum' 
- 'manhattan'
- 'canberra', 
- 'binary',
- 'minkowski' (see more in [p](#p))

### intended clusterings
```{r fig.width = 4}
viewClusters(maximum_clustered_data)
viewClusters(manhattan_clustered_data)
```

### euclidean
```{r metric_examples_euclidean, results = FALSE, fig.width = 4}
kMedoids(maximum_cluster_data, K=2, distance_method = 'euclidean')
kMedoids(manhattan_cluster_data, K=2, distance_method = 'euclidean')
```

### maximum
```{r metric_examples_maximum, results = FALSE, fig.width = 4}
kMedoids(maximum_cluster_data, K=2, distance_method = 'maximum')
kMedoids(manhattan_cluster_data, K=2, distance_method = 'maximum')
```

### manhattan
```{r metric_examples_manhattan, results = FALSE, fig.width = 4}
kMedoids(maximum_cluster_data, K=2, distance_method = 'manhattan')
kMedoids(manhattan_cluster_data, K=2, distance_method = 'manhattan')
```

## p {.tabset}
If you chose `'minkowski'` for the parameter [distance_method](#distance_method) then this is where you can input the value for the p-metric/minkowski-metric. The differences are subtle but existent, especially in the clustering functions as seen below.
Will be ignored, if other distances are chosen.


```{r full_data_def, include=FALSE}
full_data <- generateNoiseData(n=4000,lower_bounds = c(0,0),upper_bounds= c(1,1))
```

### p = 1 (manhattan)
```{r p_examples_1}
f <- kMedoids(data_5,K=4, distance_method = 'minkowski', p=1)$clustering_function
viewClusters(full_data,f)
```

### p = 1.5 
```{r p_examples_1_5}
f <- kMedoids(data_5,K=4, distance_method = 'minkowski', p=1.5)$clustering_function
viewClusters(full_data,f)
```

### p = 2 (euclidean)
```{r p_examples_2}
f <- kMedoids(data_5,K=4, distance_method = 'minkowski', p=2)$clustering_function
viewClusters(full_data,f)
```

### p = 4 
```{r p_examples_4}
f <- kMedoids(data_5,K=4, distance_method = 'minkowski', p=4)$clustering_function
viewClusters(full_data,f)
```

### p = 8 
```{r p_examples_8}
f <- kMedoids(data_5,K=4, distance = 'minkowski', p=8)$clustering_function
viewClusters(full_data,f)
```

## custom_distance_function
You can also choose a custom distance function. This distance function must be a function with the following properties. If a custom distance function is provided, the parameters [distance_method](#distance_method) and [p](#p) are ignored. 

But please consider:

- Your custom distance function must applicable on the data

- **Using an unoptimized custom distance function can slow** the calculation of kMedoids down significantly!


Inputs:   

- x and y : Two of the data row type. (The internally chosen names do not matter)

Output: 

- Must be greater or equal to 0 (semi-definite)
- Must be zero, if x and y are identical
- Must have the same output, when x and y are switched. (symmetry)


Example:
```{r custom_distance_function, result=FALSE}
discrete_distance <- function(x,y) !identical(x,y) |> as.numeric()
kMedoids(data_1, K=4, custom_distance_function = discrete_distance)
```

Because this is a main feature of the package, here is another example:
```{r globe_example_distance, result=FALSE}
# the custom function for distance on a globe
distanceByLongitudeAndLatitude <- function(x,y){
  x <- unlist(x,use.names = FALSE)[2:1] * 2*pi/360
  y <- unlist(y,use.names = FALSE)[2:1] * 2*pi/360
  round(2*6371000 * asin(sqrt(sin((y[1]-x[1])/2)^2 +
                       cos(x[1]) * cos(y[1]) *
                     sin((y[2]-x[2])/2)^2)),2)
}
```


```{r globe_example_1, echo=-c(1,6), result=FALSE}
if(require('maps')){
# load some data of word cities
slice_of_the_world <- tibble::tibble(dplyr::slice_sample(n=1000,maps::world.cities[,c('long','lat')]))

clustering <- kMedoids(slice_of_the_world, K=7, custom_distance_function = distanceByLongitudeAndLatitude)
clustering
} 
```

```{r globe_example_2, echo=-c(1,3), result=FALSE}
if(require('maps')){
viewClusters(generateNoiseData(n=4000,
                               lower_bounds = c(-180,-90), 
                               upper_bounds=c(180,90)),
             clustering$clustering_function)
} 
```


## .print_info
A logical of length 1
This is useful for debugging for example when using [custom functions](#custom_function). If TRUE some extra info will be displayed in the console

Example: 
```{r .print_info_example, fig.show='hide'}
kMedoids(data_1,K=7, .print_info = TRUE)
```

# Output
The output is a clustering object:
A list of the class clustering with at least an element called 'clustered_data'.

A clustering generated by K-Medoids has the following properties (elements of the list):

## clustered_data
The original data with a column called 'cluster' added.

```{r clustered_data}
kMedoids(data_1,K=4)$clustered_data
```

## clustering_function
A function which can relate any data point to their cluster. 
Why this is special with K-Medoids: **This clustering function CAN be applied to new data!**

```{r clustering_function_example}
training_data <- generateClusterData(n=100,
                                     clusters_mean = tibble::tibble('X'=c(0.4,0.4,0.8,1),
                                                                    'Y'=c(0.7,0.1,0.7,1)),
                                     clusters_sd=c(0.01,0.02,0.01,0.05),
                                     cluster_amount = 4)
unknown_data  <- generateClusterData(n=1000,
                                     clusters_mean = tibble::tibble('X'=c(0.4,0.4,0.8,1),
                                                                    'Y'=c(0.7,0.1,0.7,1)),
                                     clusters_sd=c(0.01,0.02,0.01,0.05),
                                     cluster_amount = 4)
more_data <- generateNoiseData(5000,
                               lower_bounds = c(0,0), 
                               upper_bounds = c(1,1))

# derive a clustering using K-Medoids
f <- kMedoids(training_data,K=4)$clustering_function

f
f(c(0.5,0.5))

# lets take a look
viewClusters(training_data,f)
viewClusters(unknown_data,f)
viewClusters(more_data,f)
```

## medoids
The clustering is calculated by the distance of the data points to the medoids of each cluster. These medoids are also part of the output.

```{r medoids_example}
kMedoids(data_1,K=4)$medoids
```

## inner_inequality
This is the sum of the distances of all data points to their cluster.
For fixed K this value can score the clustering: The lower, the better.
Be Aware: This is not enough, to directly* find an optimal value for K, as this sum will always decrease for K -> n.
* Although the inner inequality  can be used in the so called 'elbow-method'. 

```{r inner_inequality_example}
kMedoids(data_1,K=4)$inner_inequality
```

## sum_of_squares
Similar to inner_inequality but the distances are squared before being summed up.

```{r sum_of_squares}
kMedoids(data_1,K=4)$sum_of_squares
```

## mean_silhouette
The mean silhouette quantifies how good the data fits into the derived clusters.
It can be valued between -1 and 1:

- near 1: A very good clustering
- near 0: A worse clustering
- exactly 0: Most likely the clustering is a of a special case i.e. only one data point. In this case the mean silhouette gives no information about how good the clustering is.
- less than 0: A wrong clustering, in the sense that there are data points who should have been assigned to other clusters.

```{r mean_silhouette_example}
kMedoids(data_1,K=4)$mean_silhouette
```




