---
title: "The K-Means Clustering Algorithm"
output: 
  rmarkdown::html_vignette:
    toc: true
    tabset: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The K-Means Clustering Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 7,
  dev = 'svg'
)
```

```{r setup, results=FALSE , message=FALSE, warning=FALSE}
library(clusterfuck)
```


Here we will go over all information needed to use the kMeans function of the clusterfuck package.

1. First we will 


# General use
This will illustrate a use case for this clustering function:

## Categorizing spherical data
```{r general_use_1, include=FALSE}
data <- generateClusterData(n=100, 
                            cluster_amount = 4, 
                            clusters_mean = tibble::tibble('X'=c(0.2,0.7,0.5,0.9),
                                                           'Y'=c(0.2,0.2,0.7,0.3)),
                            clusters_sd = c(0.05,0.01,0.1,0.05)
                            )
```

We are given data which we want do divide into Categories, so called clusters.
```{r general_use_2}
viewData(data)
```

From insider information we are informed:
- The clusters are roughly (euclidean) spherical.
- There are supposed to be 4 clusters in this data. (note that for the Kmeans clustering algorithm needs to be known in advance.)

**Apply the K-means-algorithm with K = 4**       
```{r general_use_3}
kMeans(data, K = 4, tries = 4)
```
Well done!

## Finding representatives
Now we are given another problem. You are tasked to found a counsel debating on which color is the best color. If we were to give every humans favorite color a seat in this counsel it would be highly inefficient. Thus we want to pick only 7 representative colors. We are provided with data gathered from people formatted as a tibble of the columns red, green and blue. 

```{r fav_col_def, include=FALSE}
favorite_colors <- generateClusterData(n=200, 
                                       cluster_amount = 20,
                                       lower_bounds = c(0,0,0),
                                       upper_bounds = c(255,255,255),
                                       clusters_sd = 5+1:20,
                                       clusters_prob = c(5,5,5,5,5,1:15),
                                       colnames = c('red','green','blue'))

favorite_colors <- favorite_colors |> dplyr::filter(red>=0,red<=255,green>=0,green<=255,blue>=0,blue<=255) |> round()

favorite_colors
```
So what shall we do? We use our trusted K-Means clustering algorithm of course!
For illustrative purposes one could argue that a distance based one the maximum norm would represent a distance of color well (as then a difference between [50,50,50] 'dark grey' and [150,150,150] 'lighter gray' would be the same as between [50,0,0] 'dark blue' and [150,0,0] 'lighter blue').
Thus we apply kMeans:

```{r clustering_fav_colors}
clustering <- kMeans(favorite_colors, K = 7, distance_method = 'maximum')
```

This clustering object that gets returned has some useful properties, including centroids which are the representatives we were looking for:

```{r color_reps}
clustering$centroids
```
```{r display_rep_colors}
ddf <- clustering$centroids
image(1:nrow(ddf), 1, as.matrix(1:nrow(ddf)), 
      col=rgb(ddf$red, ddf$green, ddf$blue,maxColorValue = 255),
      xlab="", ylab = "", xaxt = "n", yaxt = "n", bty = "n")
```


## Categorizing unknown data
We continoue the example from above:

- Hey, you say, I was'nt asked what my favorite color is!
Luckily this is another example where kMeans excells at. Unknown data, on which the clustering has'nt been trained can still be clustered.

Another part of a K-Means-clustering object is the clustering function. A function which can relate a new data point to a cluster it would be part of.
Whats your favorite color? - mines a golden yellow: [255, 223, 0]

```{r adams_favorite_color}
adams_favorite_color <- c(250,180,0)
clustering$clustering_function(adams_favorite_color)
```
This is the cluster which my favorite color would be assigned to. - But what does it mean? Nothing much without context. Remember the representatives from before? Now we can defer which member of the color counsel would represent you best:

```{r favorite_representative}
clustering$centroids[
clustering$clustering_function(adams_favorite_color),]
```
```{r comparing_favorites}
ddf <- clustering$centroids[
clustering$clustering_function(adams_favorite_color),]
ddf[2,1] <- adams_favorite_color[1]
ddf[2,2] <- adams_favorite_color[2]
ddf[2,3] <- adams_favorite_color[3]
image(1:nrow(ddf), 1, as.matrix(1:nrow(ddf)), 
      col=rgb(ddf$red, ddf$green, ddf$blue,maxColorValue = 255),
      xlab="", ylab = "", xaxt = "n", yaxt = "n", bty = "n")
```
## Unknown Cluster amount 
Sometimes one may not know the cluster amount right away. Instinctively this may deter you from using K-Means to cluster. But some methods allow you to search for the 'right' cluster amount, by clustering for many different values for K and then comparing the clusterings. One of those methods is the elbow method. This makes use of the fact that this implementation of the K-Means-Clustering algorithm already provides a score of the clustering in its output. From the example above:
```{r general_use_sum_of_squares}
clustering$sum_of_squares
```
This number only makes sense in context. See this rudementary implementation of the elbow method:
```{r elbow_function_def}
findClusterAmountElbow <- function(data, check_min = 1){
  sum_of_squares <- numeric()
  improvement <- list(Inf)
  K <- 1

  sum_of_squares[[1]] <- kMeans(data,1,tries = 10)$sum_of_squares

  while(improvement[[K]]>1000 || K <= check_min){
    K <- K+1

    sum_of_squares[[K]] <- kMeans(data,K,tries = 10)$sum_of_squares

    improvement[[K]] <- sum_of_squares[[K-1]] - sum_of_squares[[K]]
  }

  improvement[improvement<1] <- Inf

  plot(1:K,sum_of_squares)
}
```

Now lets try it on the data set from above
```{r}
findClusterAmountElbow(favorite_colors,check_min = 10)
```
Here we look for an 'elbow' in the plot. Any corners in this graph can be considered a good amount of clusters in this data set. 

Something similar can be done using the silhouette method and the mean_silhouette property of the clustering object:
```{r genral_use_mean_silhouette}
clustering$mean_silhouette
```



# Parameters

## data
This is the main parameter to be provided and the data upon which the clustering algorithm will be deployed. It is recommended your data is a tibble with every row representing a data point. 

- Must have at least 1 data point.
- **`mean()` must be applicable** on the types which represent a data point. If this is an exclusion criterion for your data set, maybe try [K-Medoids](K-Medoids.html).

Here some examples: 
```{r data_definitions}
data_1 <- generateClusterData(n = 100, cluster_amount = 7)
data_2 <- generateClusterData(n = 1)
```

The data could even have multiple types, although `mean()` has to still make sense on it.


Lets go clustering!
```{r data_examples}
kMeans(data_1,K=3)
kMeans(data_2,K=1)
```

## K
The number of clusters the algorithm tries to fit the data in.
This number must be
- a positive integer 
- larger than 0
- at maximum equal to the amount of unique data points in [`data`](#data)

Some examples:
```{r K_examples}
kMeans(data_1,K=3)
kMeans(data_1,K=5)
kMeans(data_1,K=10)
kMeans(data_2,K=1)
# kMeans(data_1,K=5) throws error
```

## distance_method
This function comes with a handful of common metrics pre made. Use this parameter to choose one of the following:

```{r full_data_def, include=FALSE}
full_data <- generateNoiseData(n=1000,lower_bounds = c(0,0),upper_bounds= c(1,1))
```


- 'euclidean' (default)
- 'maximum' 
- 'manhattan'
- 'canberra', 
- 'binary',
- 'minkowski' (see more in 2.4)

### Examples {.tabset}
#### intended clusterings
```{r distance_method_examples_intended_clusterings}
viewClusters(euclidean_clustered_data)
viewClusters(maximum_clustered_data)
viewClusters(manhattan_clustered_data)
```

#### euclidean
```{r metric_examples_euclidean}
kMeans(euclidean_cluster_data, K=3, distance_method = 'euclidean')
kMeans(maximum_cluster_data, K=2, distance_method = 'euclidean')
kMeans(manhattan_cluster_data, K=2, distance_method = 'euclidean')
```

#### maximum
```{r metric_examples_maximum}
kMeans(euclidean_cluster_data, K=3, distance_method = 'maximum')
kMeans(maximum_cluster_data, K=2, distance_method = 'maximum')
kMeans(manhattan_cluster_data, K=2, distance_method = 'maximum')
```

#### manhattan
```{r metric_examples_manhattan}
kMeans(euclidean_cluster_data, K=3, distance_method = 'manhattan')
kMeans(maximum_cluster_data, K=2, distance_method = 'manhattan')
kMeans(manhattan_cluster_data, K=2, distance_method = 'manhattan')
```

## p
If you chose `'minkowski'` for the parameter `distance_method` (2.3) then this is where you can input the value for the minkowski metric. Will be ignored, if other distance methods are chosen.

### Examples {.tabset}

#### p = 1 (manhattan)
```{r p_examples_1}
set.seed(123)
p_clustering_1 <- kMeans(data_1,K=4, distance_method = 'minkowski', p=1 , tries=10)
viewClusters(full_data,p_clustering_1$clustering_function)
```

#### p = 1.5 
```{r p_examples_1_5}
set.seed(123)
p_clustering_1_5 <- kMeans(data_1,K=4, distance_method = 'minkowski', p=1.5 , tries=10)
viewClusters(full_data,p_clustering_1_5$clustering_function)
```

#### p = 2 (euclidean)
```{r p_examples_2}
set.seed(123)
p_clustering_2 <- kMeans(data_1,K=4, distance_method = 'minkowski', p=2 , tries=10)
viewClusters(full_data,p_clustering_2$clustering_function)
```

#### p = 4 
```{r p_examples_4}
set.seed(123)
p_clustering_4 <- kMeans(data_1,K=4, distance_method = 'minkowski', p=4 , tries=10)
viewClusters(full_data,p_clustering_4$clustering_function)
```

#### p = 8 
```{r p_examples_8}
set.seed(123)
p_clustering_8 <- kMeans(data_1,K=4, distance = 'minkowski', p=8 , tries=10)
viewClusters(full_data,p_clustering_8$clustering_function)
```

## custom_distance_function
You can also choose a custom distance function. This distance function must be a function with the following properties. If a custom distance is provided, the parameters `distance` (2.3) and `p` (2.4) are ignored. 

But please consider:
- Your custom metric must applicable on the data (on which `mean()` must be applicable)
- Non continous custom metrics can cause loops
- **Using an unoptimized custom metric can slow** the calculation of kMeans down significantly! 
- (For unusual custom metrics consider using [kMedoids](K-Medoid.Rmd) instead)


Inputs:   
- `x` and `y` : Two of the data row type. (The internally chosen names do not matter)

Output: 
- Must be greater or equal to 0 (semi-definite)
- Must be zero, if `x` and `y` are identical
- Must have the same output, when the values of `x` and `y` are switched. (symmetry)


Example:
```{r custom_distance_function_example_1}
discrete_distance <- function(x,y) !identical(x,y) |> as.numeric()
kMeans(data_1, K=4, custom_distance_function = discrete_distance)
```

Because this is a main feature of the package, here is another example:
```{r custom_distance_function_example_2}
WIDE_euclidean <- function(x,y) {
  x[[1]] <- 0.5*x[[1]]
  y[[1]] <- 0.5*y[[1]]
  euclidean(x,y)
  }
kMeans(data_1, K=4, custom_distance_function = WIDE_euclidean)
```

## tries
K-Means depends heavily on the starting centroids. You can account for bad luck by increasing the amount of times the algorithm is chooses new starting centroids. The clustering with minimal Inner Inequality will be returned.

Example:
```{r tries_example}
data_5 <- generateClusterData(n=1000, cluster_amount = 7, dim = 3)
kMeans(data_5,K=7,tries = 1)$inner_inequality
kMeans(data_5,K=7,tries = 10)$inner_inequality
```


## .print_info
A logical of length 1
This is useful for debugging or finding loops caused by custom metrics (2.5). If `TRUE` some extra info will be displayed in the console

Example: 
```{r .print_info_example, fig.show='hide'}
set.seed(12345)
kMeans(data_1,K=7,tries = 50, .print_info = TRUE)
```

# Output
The output is a clustering object:
A list of the class clustering with at least an element called `'clustered_data'`.
Reed more about clustering objects here TODO

A clustering generated by K-Means also has the following properties (elements of the list):

## clustered_data
The original data with a column called `'cluster'` added.

```{r clustered_data}
kMeans(data_1,K=4)$clustered_data
```

## clustering_function
A function which can relate any data point to their cluster. 
Why this is special with K-Means: **This clustering function CAN be applied to new data!**

```{r clustering_function}
clusters <- list(tibble::tibble(X=0.05,Y=0.05),
                 tibble::tibble(X=0.03,Y=0.02),
                 tibble::tibble(X=0.03,Y=0.08),
                 tibble::tibble(X=0.07,Y=0.04))
training_data <- generateClusterDataFromPaths(n=50, clusters)
unknown_data <- generateClusterDataFromPaths(n=100, clusters)
more_data <- full_data <- generateNoiseData(n=5000,lower_bounds = c(0,0),upper_bounds= c(0.1,0.1))

# derive a clustering using K-Means
f <- kMeans(training_data,K=4)$clustering_function

f
f(c(0.5,0.5))

# lets take a look
viewClusters(training_data,f)
viewClusters(unknown_data,f)
viewClusters(more_data,f)
```

## centroids
The clustering is calculated by the distance of the data points to the centroids (means) of each cluster. These centroids are also part of the output.

```{r centroids}
kMeans(data_1,K=4)$centroids
```

## inner_inequality
This is the sum of the distances of all data points to their cluster.
For fixed K this value can score the clustering: The lower, the better.
Be Aware: This is not enough, to directly* find an optimal value for `K`, as this sum will always decrease for K \rightarrow n.
* Although the inner inequality  can be used in the so called 'elbow-method'. More here TODO

```{r inner_inequality}
kMeans(data_1,K=4)$inner_inequality
```

## sum_of_squares
Similar to `inner_inequality` (3.4) but the distances are squared before being summed up.

```{r sum_of_squares}
kMeans(data_1,K=4)$sum_of_squares
```

## mean_silhouette
The mean silhouette quantifies how good the data fits into the derived clusters.
It can be valued between -1 and 1:

- near 1: A very good clustering
- near 0: A worse clustering
- exactly 0: Most likely the clustering is a of a special case i.e. only one data point. In this case the mean silhouette gives no information about how good the clustering is.
- less than 0: A wrong clustering, in the sense that there are data points who should have been assigned to other clusters.

```{r mean_silhouette}
kMeans(data_1,K=4)$mean_silhouette
```
