  ---
title: "The OPTICS Clustering Algorithm"
output: 
  rmarkdown::html_vignette:
    toc: true
    tabset: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The OPTICS Clustering Algortihm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 7,
  dev='svg'
)
```

```{r setup, results=FALSE , message=FALSE, warning=FALSE}
library(clusterfuck)
```

# The OPTICS Clustering Algortihm
This is an overview about the OPTICS (Ordering points to identify the clustering structure Clustering algorithm of the clusterfuck package.
OPTICS is a density-based clustering algorithm similar to DBSCAN, with the main difference that OPTICS returns an ordering of the data.

OPTICS does not directly compute clusters. Instead it produces an ordering of the data points together with reachability distances that describe the density structure of the data.

## 1. General use
We start with creating some test data
```{r general_use_1}
data <- generateClusterData(n=300)
```

This is what it looks like:
```{r general_use_2}
viewData(data)
```

**Apply the OPTICS algorithm with a neighborhood of 0.2 and a minimal number of points for a cluster to be formed 3**
```{r general_use_3}
result <- optics(data, epsilon=0.2, min_Pts=3)
```

## 2. Parameters

### 2.1 data (required)
This is the main parameter to be provided and the data upon which the clustering algorithm will be deployed. It is recommended your data is a tibble with every row representing a data point. data must have at least 1 data point having at least 1 dimension. 

Here some examples: 
```{r data_examples_1}
data_1 <- optics_wikipedia_data
data_2 <- generateClusterData(n=2000)
data_3 <- generateClusterData(n=100)
```

We can cluster this, and visualize the result:
```{r data_examples_2}
result_1 <- optics(data_1, epsilon=0.2, min_Pts=3)
cluster <- extractClusters(result_1, epsilon = 0.1)
data_1$cluster <- cluster
viewClusters(data_1)

result_2 <- optics(data_2, epsilon=0.2, min_Pts=3)
cluster <- extractClusters(result_2, epsilon = 0.1)
data_2$cluster <- cluster
viewClusters(data_2)

result_3 <- optics(data_3, epsilon=0.2, min_Pts=3)
cluster <- extractClusters(result_3, epsilon = 0.1)
data_1$cluster <- cluster
viewClusters(data_3)
```
Points assigned to cluster 0 represent noise points that were not
assigned to any cluster.

### 2.2 epsilon
The maximum radius used to determine neighbors of a point. In OPTICS, epsilon mainly acts as an upper bound radius for the neighborhood. Clusters can later be extracted using a smaller radius.
epsilon must be a positive number greater than 0.

This can be be observed when clustering the same data with different epsilon values:
```{r epsilon_examples_1}
result_small_epsilon <- optics(data_1, epsilon=0.1, min_Pts=5)
viewClusters(result_small_epsilon$clustered_data)
result_medium_epsilon <- optics(data_1, epsilon=0.2, min_Pts=5)
viewClusters(result_medium_epsilon$clustered_data)
result_big_epsilon <- optics(data_1, epsilon=0.4, min_Pts=5)
viewClusters(result_big_epsilon$clustered_data)
```

### 2.3 min_Pts
The number of points that have to be within epsilon neighborhood of each other to compute a core distance. Points with less than min_Pts neighbors within epsilon distance are not considered core points.
min_Pts must be a positive integer greater than 0. 

This can be be observed when clustering the same data with different min_Pts values:
```{r minPts_examples_1}
result_small_minPts <- optics(data_1, epsilon=0.2, min_Pts=3)
viewClusters(result_small_minPts$clustered_data)
result_medium_minPts <- optics(data_1, epsilon=0.2, min_Pts=5)
viewClusters(result_medium_minPts$clustered_data)
result_big_minPts <- optics(data_1, epsilon=0.2, min_Pts=10)
viewClusters(result_big_minPts$clustered_data)
```
## 3. Output
The output is a clustering object:
A list of the class "clustering" and "optics", which is required for the reachability plot.

### 3.1 order
The order in which the points were processed.
```{r order}
result$order
```

### 3.2 reachdist
The distance describing how densely each point is connected to previously processed points.
```{r reachdist_example}
result$reachdist
```

### 3.3 coredist
The core distance of each point.
```{r coredist_example}
result$coredist
```

## 3.4 data
The original dataset used for computing the OPTICS ordering.
```{r data_example_1}
result$data
```
data can be combined with the cluster list that is computed with extractClusters() to obtain a table as clustered_data:
```{r data_example_2}
extractClusters(result, epsilon = 0.2 )
```

## 3.5 minPts
The original value of min_Pts used when running the OPTICS algorithm.
```{r min_Pts_example}
result$minPts
```
When clusters are extracted using extractClusters(), clusters with fewer
points than minPts are treated as noise.

## 4. Extracting Clusters
After running OPTICS, we can extract the clusters at a specific epsilon threshold to get a flat clustering, similar to DBSCAN. 
```{r extracting_Clusters}
cluster<- extractClusters(result, epsilon=0.1)
data$cluster <- cluster
```
Now, we can use viewClusters() on data to plot the clustered data:
```{r extracting_Clusters_viewClusters}
viewClusters(data)
```

## 5. Reachability Plot
The reachability plot is the main visualization tool for OPTICS.
```{r reachability_example}
plot(result, epsilon=0.1)
```
Valleys in the reachability plot correspond to clusters of points with similar density. Each colour represents a cluster.

## 6. optics_wikipedia_data 
The wikipedia article of the OPTICS algorithm displays a example of OPTICS clustering on a certain dataset. This is recreated in data-raw/optics_wikipedia_data.R
The dataset is build from three paths, vaguely representing three clusters, and additional noise data over the paths.

### 6.1 sample_points_from_path
Generate twodimensional noise points between 0 and 1.

#### 6.1.1 Parameters
path_fun is the function describing the paths that should be turned into data points.
n is the number of wanted points to be generated along each path.
sd is the standard noise deviation on the paths.

#### 6.1.2 Output
A tibble of points vaguely representing the input path(s).


### 6.2 generate_noise
Generate twodimensional noise points between 0 and 1.

#### 6.2.1 Parameter
n is the number of wanted noise points.

#### 6.2.2 Output
A tibble of randomly distributed points between 0 and 1.


### 6.3 generate_optics_test_data
Generates clustered data along given paths with Gaussian noise.

#### 6.3.1 Parameters
paths is the list of paths to be turned into datapoints.
n is the number of wanted points to be generated along each path.
n is the number of wanted number of noise points over the data.
sd is the standard noise deviation on the paths, 0.035 in this wikipedia data example.

#### 6.3.2 Output
A tibble representing the paths + noise as data points.

### 6.4 Recreating the example
```{r wikipedia_example}
path1 <- tibble::tibble(
  X = c(0.2,0.3,0.4,0.3,0.2),
  Y = c(0.3,0.4,0.3,0.2,0.3)
)
path2 <- tibble::tibble(
  X = c(0.45,0.5,0.45,0.4,0.45),
  Y = c(0.9,0.85,0.8,0.85,0.9)
)
path3 <- tibble::tibble(
  X = c(0.85,0.8,0.93,0.85),
  Y = c(0.5,0.28,0.28,0.5)
)
paths <- list(path1, path2, path3)
optics_wikipedia_data <- generate_optics_test_data(paths, n_points = 500, noise_points = 100, sd = 0.035 )
wikipedia_result <- optics(optics_wikipedia_data, epsilon=0.15, min_Pts=3)
```
This can then be visualized as shown before, and we see a result similar to the example on the OPTICS wikipedia article.
```{r wikipedia_visualized}
wikipedia_cluster <- extractClusters(wikipedia_result, epsilon=0.1)
optics_wikipedia_data$cluster <- wikipedia_cluster
viewClusters(optics_wikipedia_data)
plot(wikipedia_result, epsilon=0.1)
```
