---
title: "The Spectral Clustering Algortihm"
output: 
  rmarkdown::html_vignette:
    toc: true
    tabset: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The Spectral Clustering Algortihm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 7,
  dev='svg'
)
```

```{r setup, results = FALSE}
library(clusterfuck)
```

Here we will go over all information needed to use the spectralClustering function of the clusterfuck package

# General use
This algorithm is perfect for data sets where

- DBSCAN and OPTICS fail because e.g. all points are connected by 'bridges'
- K-Means and K-Medoids fail, because the clusters are non spherical. 

One such data set is this one:
```{r general_use_viewData}
three_connected_concentric_circles |> viewData()
```

**Apply the Spectral-clustering-algorithm**       
```{r general_use_apply_clustering}
three_connected_concentric_circles |>
spectralClustering(k=3,
                   gamma = 50,
                   cluster_algorithm = 'K-Medoids',
                   K = 3
                   )
```





# Parameters

## data
This is the main parameter to be provided and the data upon which the clustering algorithm will be deployed. It is recommended your data is a tibble with every row representing a data point. Must have at least 1 data point. 

Here some examples: 
```{r data_examples_1}
data_1 <- generateClusterData(n=100)
data_2 <- generateClusterData(n=1)
data_3 <- study_courses_data
data_5 <- generateClusterData(n=100,
                              clusters_mean = tibble::tibble('X'=c(0.4,0.4,0.8,1),
                                                             'Y'=c(0.7,0.1,0.7,1)),
                              clusters_sd=c(0.01,0.02,0.01,0.05),
                              cluster_amount = 4)
```

The data can even have multiple types
```{r data_examples_2}
numbers_and_characters
```

```{r data_examples_3, include=FALSE}
numericAndCharacterMixedDistance <- function(x,y){
 return(length(x[[2]])-length(y[[2]]) +  abs(x[[1]]-y[[1]]))
}
```


Lets go clustering!
```{r data_examples_4, results = FALSE}
spectralClustering(data_1,k=3,gamma=50,K=3)
spectralClustering(data_3,k=1,gamma=50,custom_distance_function = study_courses_distance,K=3)
spectralClustering(numbers_and_characters,k=3,gamma=50,custom_distance_function = numericAndCharacterMixedDistance,K=3)
```

## k
The Dimension of the space the data points get projected to. 
A higher amount of k can decrease the dependency on choosing the right value for 
[gamma](#gamma) **significantly**!

Some examples:
```{r k_examples, results = FALSE}
spectralClustering(data_1,k=1,gamma=50,K=3)
spectralClustering(data_1,k=2,gamma=50,K=3)
spectralClustering(data_1,k=3,gamma=50,K=3)
```

## mercer_kernel 

The mercer kernel is a measure of similarity between data points and is used in the spectral projection part of the algorithm.

Here You can select one of

- `'gauss'`: $\exp{(-\gamma~~||x-y||_2)}$ (default)
If chosen please provide a non negative value for [gamma](#gamma)  
  (Note: in this implementation as we use the stats::dist function, we do not square the distances as it would be another calculation) 
- `'linear'`: $x^Ty$ 

```{r, results=FALSE}
spectralClustering(three_connected_concentric_circles,
          k=3,
          mercer_kernel='gauss',
          gamma=1,
          K=3)
```
 


## gamma {.tabset}
This is one of the many parameters ([gamma](#gamma), [kernel_epsilon](#epsilon), [mercer_kernel](#mercer_kernel), [custom_mercer_kernel](#custom_mercer_kernel), [distance_method](#distance_method), [p](#p) and [custom_distance_function](#custom_distance_function)) used to select the kernel.

The value of gamma describes how much the distances will be exaggerated by the projection. This is very finicky to get right. Thus we recommend choosing a higher value for [k](#k) which can decrease the dependency on choosing the right value for gamma.

Will be ignored, if 

- another [mercer_kernel](#mercer_kernel) or
- a [custom_mercer_kernel](#custom_mercer_kernel) got selected

### low gamma 
```{r gamma_example_low, results=FALSE}
spectralClustering(three_connected_concentric_circles,
                   k=3,
                   gamma=1,
                   K=3,
                   tries=10)
```

### good gamma 
```{r gamma_example_good, results=FALSE}
spectralClustering(three_connected_concentric_circles,
                   k=3,
                   gamma=50,
                   K=3,
                   tries=10)
```

### high gamma
```{r gamma_example_high, results=FALSE}
spectralClustering(three_connected_concentric_circles,
                   k=3,
                   gamma=100,
                   K=3,
                   tries=10)
```

## kernel_epsilon {.tabset}
This is one of the many parameters ([gamma](#gamma), [kernel_epsilon](#epsilon), [mercer_kernel](#mercer_kernel), [custom_mercer_kernel](#custom_mercer_kernel), [distance_method](#distance_method), [p](#p) and [custom_distance_function](#custom_distance_function)) used to select the kernel.

Its value represents a distance limit at which the similarity of data points will be set to `0`.

Must be a numeric greater than `0`.

### low epsilon
```{r kernel_epsilon_1, results=FALSE}
spectralClustering(two_connected_concentric_circles,
                   k=3,
                   gamma=50,
                   kernel_epsilon = 0.01,
                   K=3)
```

### high epsilon
```{r kernel_epsilon_2, results=FALSE}
spectralClustering(two_connected_concentric_circles,
                   k=3,
                   gamma=50,
                   kernel_epsilon = 0.5,
                   K=3)
```

## custom_mercer_kernel
This is one of the many parameters ([gamma](#gamma), [kernel_epsilon](#epsilon), [mercer_kernel](#mercer_kernel), [custom_mercer_kernel](#custom_mercer_kernel), [distance_method](#distance_method), [p](#p) and [custom_distance_function](#custom_distance_function)) used to select the kernel.

Using this parameter you can provide a custom mercer kernel to be used in the spectral clustering algorithm.

If this is provided, any other aforementioned parameter will be ignored.

```{r custom_mercer_kernel, result=FALSE}
discrete_similarity <- function(x,y) identical(x,y) |> as.numeric()
spectralClustering(tibble::tibble(X=c(1,1,4,1,5,1,4)),
                   k=2,
                   custom_mercer_kernel = discrete_similarity,
                   K=3,
                   tries=10)
```

## distance_method
This is one of the many parameters ([gamma](#gamma), [kernel_epsilon](#epsilon), [mercer_kernel](#mercer_kernel), [custom_mercer_kernel](#custom_mercer_kernel), [distance_method](#distance_method), [p](#p) and [custom_distance_function](#custom_distance_function)) used to select the kernel.

It is a use friendly way to use common distance functions in spectral clusterings.
Calculates the gaussian kernel in [mercer_kernel](#mercer_kernel) using this distance instead of the euclidean distance. This means: please provide a non negative value for [gamma](#gamma)

Choose one of:

- `'euclidean'` (default)
- `'maximum'` 
- `'manhattan'`
- `'canberra'` 
- `'binary'`
- `'minkowski'` (see more in [p](#p))

```{r distance_method_example, results=FALSE}
spectralClustering(maximum_cluster_data,
                   k=1,
                   gamma=10,
                   distance_method = 'maximum', 
                   K=2)
```

## p 
This is one of the many parameters ([gamma](#gamma), [kernel_epsilon](#epsilon), [mercer_kernel](#mercer_kernel), [custom_mercer_kernel](#custom_mercer_kernel), [distance_method](#distance_method), [p](#p) and [custom_distance_function](#custom_distance_function)) used to select the kernel.

If you chose `'minkowski'` for the parameter [distance_method](#distance_method) then this is where you can input the value for the p-metric/minkowski-metric. 
```{r p_example, results=FALSE}
spectralClustering(manhattan_cluster_data,
                   k=1,
                   gamma=10,
                   distance_method = 'minkowski',
                   p = 1.3,
                   K=2)
```


## custom_distance_function
You can also choose a custom distance function. This distance function must be a function with the following properties. If a custom distance function is provided, the parameters [distance_method](#distance_method) and [p](#p) are ignored. 

But please consider:

- Your custom distance function must applicable on the data
- **Using an unoptimized custom distance function can slow** the calculation down significantly!


Inputs:   

- x and y : Two of the data row type. (The internally chosen names do not matter)

Output: 

- Must be greater or equal to 0 (semi-definite)
- Must be zero, if x and y are identical
- Must have the same output, when x and y are switched. (symmetry)


Example:
Clustering study courses by projecting them to the 2D space
```{r custom_distance_function, result=FALSE}
spectralClustering(study_courses_data, 
                   k=2, 
                   gamma = 1,
                   custom_distance_function = study_courses_distance,
                   K=7)
```

Because this is a main feature of the package, here is another example:
Clustering the WORLD
```{r globe_example_distance, result=FALSE}
# the custom function for distance on a globe
distanceByLongitudeAndLatitude <- function(x,y){
  x <- unlist(x,use.names = FALSE)[2:1] * 2*pi/360
  y <- unlist(y,use.names = FALSE)[2:1] * 2*pi/360
  round(2*6371000 * asin(sqrt(sin((y[1]-x[1])/2)^2 +
                       cos(x[1]) * cos(y[1]) *
                     sin((y[2]-x[2])/2)^2)),2)
}
```


```{r globe_example_1, echo=-c(1,6), result=FALSE}
if(require('maps')){
# load some data of word cities
slice_of_the_world <- tibble::tibble(dplyr::slice_sample(n=1000,
                                                         maps::world.cities[,c('long','lat')]))

spectralClustering(slice_of_the_world, 
                   k=3, 
                   gamma = 0.00001,
                   custom_distance_function = distanceByLongitudeAndLatitude,
                   K=7)
} 
```

## cluster_algorithm
One of the **most important** parameters of this clustering algorithm. This algorithm will be used to cluster the projected data points.

Choose one of:

- `'K-Means'` (default) [see vignette](K-Means.html)
- `'K-Medoids'` [see vignette](K-Medoids.html)
- `'hierarchical clustering'` [see vignette](hierarchical-clustering.html) 
- `'DBSCAN'` [see vignette](dbscan.html)
- `'OPTICS'` [see vignette](optics.html)

Mote that you need to provide all necessary parameters for the selected clustering algorithm using the `...` parameter.

```{r cluster_algorithm_example}
spectralClustering(two_concentric_circles, 
                   k=1, 
                   gamma = 20,
                   cluster_algorithm = 'DBSCAN',
                   epsilon=0.0001, 
                   min_Pts=5
                   )
```
## ...
Arguments to be passed on to the [selected clustering algorithm](#cluster_alorithm)

## .print_info
A logical of length 1.
If TRUE some extra info will be displayed in the console

Example: 
```{r .print_info_example, fig.show='hide'}
spectralClustering(data_1,k=5,gamma=30,K=7, .print_info = TRUE)
```

# Output
The output depends on the chosen clustering algorithm

Most often a clustering object:
A list of the class clustering with at least an element called 'clustered_data'.

A clustering generated like this has the following properties (elements of the list):

## clustered_data
The original data with a column called 'cluster' added.

```{r clustered_data}
spectralClustering(data_1,k=5,gamma=30,K=7)$clustered_data
```

## clustered_projected_data

The projected data with a column called 'cluster' added.

```{r clustered_projected_data}
spectralClustering(data_1,k=5,gamma=30,K=7)$projected_clustered_data
```

## more
Further Elements depending on the clustering algorithm,
